---
title: "Statistical Rethinking Notes - Chapter 2"
author: Zachary Levonian
date: 2022
output: pdf_document
---

# Chapter 2

Notes on chapter 2.

Bayesian data analysis:
For each possible explanation of the data,
Count all the ways the data can happen.
Explanations with more ways to produce the data are more plausible.

$\text{Pr}(W, L|p) = \frac{(W + L)!}{W!L!}p^W(1-p)^L$ where $W$ is the number of water hits and $L$ is the number of land hits.

```{r}
dbinom(6, 9, 0.7)
```

```{r}
plot(1:100, dbinom(1:100, 100, 0.7))
```

```{r}
xs <- seq(0, 1, 0.001)
plot(xs, dbinom(9, 9, xs), type="l", col="blue", 
     main="Binomial density plot for different number of hits (of 9 total)")
lines(xs, dbinom(5, 9, xs), col="green")
lines(xs, dbinom(6, 9, xs), col="red")
lines(xs, dbinom(7, 9, xs), col="gray")
lines(xs, dbinom(8, 9, xs), col="black")
legend(0, 1, 
       legend=c("5 hits", "6 hits", "7 hits", "8 hits", "9 hits"),
       col=c("green", "red", "gray", "black", "blue"), 
       lty=1, cex=0.8,
       box.lty=0)

```

Question 1: why is it okay to set the prior to 1 (rather than 1 / sum(prior))? (in the code example given in the lecture) (answer: because we will normalize after anyway, so it doesn't matter.)

Question 2: why is the evidence of a single W or L a line (and not some other shape)?


